Depression and suicidality as costly signals of need

### 2021-04-02 Paper.rmd
Working on intro, added refs to bib file

Durkheim’s pioneering study of suicide [@durkheim1897] arguably launched sociology as a quantitative discipline. Sociological investigations of suicide flourished through much of the 20th century, but began to be supplanted by medical and public health approaches as the biomedical model of mental health took hold in the late 1970's and early 1980's [@wray2011; @syme2020]. The biomedical model tends to frame suicide as an individiual-level problem, e.g., as one involving psychiatric disturbance, drug addiction, or severe stress [@wray2011]. This theoretical approach has made little progress: Despite substantial investment by funding agencies, rates of suicide have remained largely unchanged over the last 100 years (confirm and cite).

Early on, anthropologists also tackled the problem of suicide as manifested in the small, traditional socities in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal (e.g., Malinowski, 1985; Niehaus, 2012). Some ethnographers, such as Hezel (1987), conceptualized suicide as a form of anger, whereas others, such as Giddens (1964), conceptualized suicide as a form of social pressure. Counts (1980) emphasized the powerlessness of suicide victims. Common to almost all theoretical and empirical investigations of suicide is a focus on *completed* suicides, i.e., suicide deaths. @SymeEtAl2016, in contrast, argued that because the vast majority of suicidal behavior does not result in death, the proper focus should be on suicide ideation and suicide attempts, not deaths.

Within anthropology, Raymond Firth was likely the first to put the focus on suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed social conflict or loss and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, and immediate reintegration into the community following negative events (Firth 1967[not original date]). Compared to suicide, however, anthropological explanations for depression are a much more recent occurrence (cite examples), despite frequent reports of depression or depression-like behavior in a wide range of societies (cite) and the potential for cross-cultural research to help understand depressive states (cite).

Alternative depression idea:
- move Firth up before Syme et al. 2016 and tweak slighlty or reword Firth and put behind Syme in the same paragraph
- delete the previous depression stuff
- then follow with this
As with suicide, research on depression has been dominated by biomedical approaches of limited success. However, evolutionary and anthropological explanations for low mood have become more common. In general, these approaches see sadness and minor depression as ways to motivate adaptive behavior changes in response to negative outcomes in one's environment (cite Nesse and others), mainly differing in how they view major depression. While some approaches present most forms of major depression as dysfunctional (inflammation and some of Nesse), others have extended the logic on the benefits of low mood and reduced behavior to more severe instances of sadness. Of the possible benefits, the ability of depression to help solve fitness-relevant in times of social conflict is the most emphasized, with depression being suggested to help by reallocating resources with the body for problem solving (cite ARH), reducing the costs of interacting with higher status individuals (cite), or serving as a means to both signal one's need and bargain for increased support (cite).

>>>>

Durkheim’s pioneering study of suicide [@durkheim1897] helped launch sociology as a quantitative discipline. Sociological investigations of suicide flourished through much of the 20th century, but were supplanted by medical and public health approaches as the biomedical model of mental health took hold in the late 1970's and early 1980's [@wray2011; @syme2020]. The biomedical model tends to frame suicide as a problem with the individual, e.g., as one involving psychiatric disturbance, drug addiction, or stress sensitivity [@wray2011]. This theoretical approach has made little progress: Rates of suicide, which vary widely by country, have remained relatively flat in countries for which there are long-term data, such as the US, UK, Japan, and Sweden [@owidsuicide].

Early on, anthropologists also tackled the problem of suicide as manifested in the small traditional societies in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [e.g., @Malinowski:1932; @niehaus2012]. Some ethnographers, such as @Hezel:1987, conceptualized suicide as a form of anger, whereas others, such as @Giddens:1964, conceptualized suicide as a form of social pressure. @Counts:1980 emphasized the powerlessness of suicide victims. Common to almost all theoretical and empirical investigations of suicide is a focus on *completed* suicides, i.e., suicide deaths. @SymeEtAl2016, in contrast, argued that because the vast majority of suicidal behavior does not result in death, the proper focus should be on suicide ideation and suicide attempts, with death as an unintended consequence. See Figure \@ref(fig:cdc).

```{r cdc, fig.width=8, fig.cap="US Suicide attempt and completion rates by age and sex (2001-2011)."}
library(hrafsuicidedata)
cdc %>% 
  pivot_longer(c(`Suicide rate`, `Attempt rate`)) %>% 
  ggplot(aes(age2, value, colour = sex, linetype = name)) + 
  geom_line() +
  hagenutils::scale_color_binary() +
  guides(colour = guide_legend(title='Sex'), linetype = guide_legend(title=NULL)) +
  labs(x = '\nAge (years)', y = 'Rate\n(per 100,000)', hjust=0) +
  theme_minimal(15) +
  theme(axis.title.y = element_text(angle = 0))
```

Within anthropology, Raymond Firth was likely the first to put the focus on suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed social conflict or loss, and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, and immediate reintegration into the community following negative events [@firth1936: @Firth:1961].

Suicidality is closely linked to depression. Suicidal ideation is one of nine diagnostic criteria for a major depressive episode [MDE, @dsm5], it is associated with depression across cultures [@haroz2017], and depression is a major risk factor for suicidal behavior [@hawton2013]. Compared to suicide, however, there are fewer anthropological investigations of depression [@kleinman1985] despite frequent reports of depression or depression-like behavior in a wide range of societies [@haroz2017; @liu2020] and the potential for cross-cultural research to help understand depressive states [@haroz2017].

Evolutionary theories of depression tend to either ignore suicidality, or take it as evidence that depression is dysfunctional [for reviews of evolutionary theories of depression, see @].

and anthropological explanations for low mood have become more common. In general, these approaches see sadness and minor depression as ways to motivate adaptive behavior changes in response to negative outcomes in one's environment (cite Nesse and others), mainly differing in how they view major depression. While some approaches present most forms of major depression as dysfunctional (inflammation and some of Nesse), others have extended the logic on the benefits of low mood and reduced behavior to more severe instances of sadness. Of the possible benefits, the ability of depression to help solve fitness-relevant in times of social conflict is the most emphasized, with depression being suggested to help by reallocating resources with the body for problem solving (cite ARH), reducing the costs of interacting with higher status individuals (cite), or serving as a means to both signal one's need and bargain for increased support (cite).

Alternative depression idea:
- move Firth up before Syme et al. 2016 and tweak slighlty or reword Firth and put behind Syme in the same paragraph
- delete the previous depression stuff
- then follow with this

As with suicide, research on depression since the 1980's has been dominated by biomedical approaches of limited success [@patten2016; @ormel2019].
